---
title: "Cleaning and Exploring Data"
author: "Jun Tan 26370084"
output: word_document
---

### Introduction

This assignment has three goals: 
(a) to familiarize you with some of the basic 
functions in R and how to use them, 
(b) to learn how to subset data frames to eliminate unusual records and values,
(c) to review/introduce some of the key
concepts in exploratory data analysis.

You will need to install the ggplot2 and RColorBrewer packages. Do this now before starting your HW. Use the packages tab in the pane where the plots are displayed to install the packages.

You will work with data published by the San Francisco Chronicle about housing sales. We begin by reading the data from the Web into R.

```{r}
load(url("http://www.stat.berkeley.edu/users/nolan/data/Projects/SFHousing.rda"))
```

This command loads the *housing* data frame into your work space. 
The following functions may be useful in answering 
the questions:
**class(), cor(), length(), dim(), ls(), names(), mean(), min(), max(), quantile(), summary().**


1. What are the names of the variables in the data frame?
What type of data are they, i.e., are they numeric, logical,
character, factors? 

```{r}
names(housing)
sapply(housing, class)
```

2. Can you confirm that there are measurements on 281,506 houses  
available in *housing*?

```{r}
dim(housing)
nrow(housing)
```


The data do not indicate the units of measurement and some of the variable names are a bit terse. The variables and their units are:

* county - County name
* city - City name
* zip - Zip code
* street - Street address
* price - Sale price in dollars
* br - Number of bedrooms
* lsqft - Size of lot in square feet
* bsqft - Size of building in square feet
* year - Year house was built
* long - Longitude of house location
* lat - Latitude of house location

We are interested in the relationship between the price per square foot  and the size of the house.  Before we study this relaionship, we 
want to examine these and other variables to check if we need to clean the data.


### Property Size

3. Examine the property (lot) size with summary statistics. What do you see?

```{r}
summary(housing)
```
The summary shows that some zip codes are missing but no counties or cities are missing. The amounts of cities and zip codes are classified as other are extremely high. Moreover, the max price is a lot higher than the 3rd quantile. 


4. An acre is 43560 sqft. It seems that houses with plots more 
than 3 to 5 acres are different types of properties than those without. 
Set *lsqft* to NA if the home has a lot of 3+ acres.

```{r}
housing$lsqft[(housing$lsqft/43560) > 3] = NA
```


5. Some of the tiniest houses measure 100-200 square feet 
(see http://www.housebeautiful.com/design-inspiration/house-tours/a5836/hikari-box-tiny-house-shelter-wise/) 
Since these are tiny, modern "eco homes", we can exclude houses that don't have a lot square footage of a couple times that. Again, set *lsqft* to NA for those homes with less than a 400 square foot lot.

```{r}
housing$lsqft[housing$bsqft < 400] = NA
```


## House Size 

6. As with lot size, there are some unusually large houses. How big are the 10 largest houses? (Use *order()* to help find this information)

```{r}
orderSize = sort(housing$bsqft)
tail(orderSize, 10)

```

7. Drop any house over 20000 square feet from the data frame.

```{r}
drop = housing$bsqft > 20000
housing = housing[!drop, ]
```


## Relationship between lot size and house size

8. How many houses have a larger building square footage than lot square footage? How can this happen?

```{r}
greater = (housing$bsqft > housing$lsqft)
amount = length(greater[greater==TRUE])
amount
```
This number shows that a lot of buildings are tall such that they can provide more living space.

9. Remove all houses from the data frame that have a building size that is more than 4 times the lot size.

```{r}
drop = housing$bsqft > (housing$lsqft * 4)
housing = housing[!drop, ]
```

## Price per square foot

10. We said that we are interested in the price per square foot of the houses. This variable is not in our data frame. Create a new variable, *ppsf*, in *housing*.

```{r}
ppsf = housing$price / housing$bsqft
housing$ppsf = ppsf
```

## Extraneous Variables

11. There are a few variables related to the quality of the data that are
not mentioned in the list above. Figure out which these are and drop them from the data frame. Use subsetting by exclusion to do this.

```{r}
ind = names(housing) != "quality" & names(housing) != "match" & names(housing) != "wk"
housing = housing[ ,ind]
```


## Date

Since the housing prices changed a lot in the time period of observation,
let's restrict our view to houses that were sold in the first two quarters of 2004.  To do this, we need to convert *date* into a  more appropriate data type. The following code creates a vector in a POSIX format and then
pulls out the month and year from each date and saves them in numeric vectors.

```{r}
month = as.numeric(format(housing$date, format="%m"))
year = as.numeric(format(housing$date, format="%Y"))
```

12. Use these vectors to create a subset of *housing* with sales from the first half of 2004.

```{r}
year_logic = !is.na(year ==2004)
month_logic = !is.na(month <=6 )
housing = housing[year_logic & month_logic, ]
```

You should have about 29,000 houses left in the data frame.

## Similar Locations

Another approach to exploring and analyzing data is to examine subsets that are similar. For example, we can `zoom in' to the east bay and examine the relationship between price per square foot and size of the house for houses sold in the following cities:

```{r}
someCities = c("Albany", "Berkeley", "El Cerrito", "Emeryville", 
               "Piedmont", "Richmond", "Lafayette", "Walnut Creek",
               "Kensington", "Alameda", "Orinda", "Moraga")
```



13. Use this character vector to take a subset of sales in those cities. You may find the operator *%in%* useful here.

```{r}
sales = housing[housing$city %in% someCities, ]
```

The *city* variable is a factor and when all of the records for a particular city have been removed from the data, the level for that factor remains as a possible level. This means that summaries and other statistics will include these `empty' levels. To drop them, we can modify
*city* as follows

```{r}
housing$city = droplevels(housing$city)
```

Note that we expect that your subset of house sales is still in the data frame called *housing*.  If not, you might want to reassign *housing* to your current data frame and remove (*rm()* ) all other intermediate and 
duplicate data frames.


## Plot

Our data should be ready to plotting. Below is plot that uses color to indicate density of points. You will learn how to make this and other plots soon. The following code chunk is set up so as to not run when knitted. You will want to remove the `eval=TRUE` from within the curly braces when you are ready to produce the final plot.

```{r, eval=TRUE}
require(ggplot2)
require(RColorBrewer)
housing$densCol = densCols(housing$bsqft, 
        y = log(housing$ppsf), 
        colramp = colorRampPalette(brewer.pal(9,"BuPu")[3:9]))

ggplot(housing) + 
  geom_point(aes(x = bsqft, y = ppsf, col = densCol), size = 1) +
  scale_color_identity() + 
  xlim(0, 6000) + ylim(0,1250) +
  xlab("Building Square Footage") + ylab("Price per Square Foot") + 
  theme_bw() 
```

Upload your completed Rmd file to bcourses. Do not upload the knitted file.

