---
title: "Lab 5"
output: html_document
---


## Introduction

In class, we wrote the following function to simulate St. Petersberg's Paradox. 

```{r}
double.num = function(numBets) {
   
  if (!is.numeric(numBets)) stop("numBets must be numeric")
  outcomes = c(-1, 1)
  
  for (i in 1:numBets) {
    winLoss = sample(outcomes, size = 1)
    if (winLoss > 0) return(i)
  }
  return(NA)
}
```


Q1. Add a check to this function for `numBets`. Our function expects that `numBets` is a single number. If it contains more numbers, then the function should use only the first and it should issue a warning to that effect. Call this revised function `double.revised`. Test that it works as expected.

```{r}
double.revised = function(numBets) {
  
  if (!is.numeric(numBets)) stop("numBets must be numeric")
  if (length(numBets) >1) {
    warning("numBets should contains only one number")
    numBets = numBets[1]
  }
  outcomes = c(-1, 1)
  
  for (i in 1:numBets) {
    winLoss = sample(outcomes, size = 1)
    if (winLoss > 0) return(i)
  }
  return(NA)
}
```

Put your tests here 
```{r, collapse = TRUE, error=TRUE}
double.revised(c(1,2,3))
double.revised(c(100, 10, 1))
```


Q2. Write a new function that takes a dollar amount rather than the number of bets as input. This should be a required parameter called `wallet`.  Your code needs to determine the maximum number of bets that can be placed given the value provided for `wallet`.  (Remember that the number of bets should be an integer). To help you, recall that 1 + 2 is 4-1 and 1 + 2 + 4 is 8 - 1  and 1 + 2 + 4 + 8 is 16 - 1.  Do you see a pattern?

This function should return the gain. Note that the gain is one of two possible values: 1 or the total amount lost. Call this function `double.gain`.


```{r}
double.gain = function(wallet) {
  numBets = floor(sqrt(wallet+1))
  return(double.revised(numBets))
}
```

Be sure to test this function.

```{r, collapse = TRUE, error=TRUE}
# Put some tests here 
double.gain(3)
double.gain(22)
```



Q3. Write a new function that can accept either the number of bets to place or the amount in the wallet or both.  The number of bets, `numBets`, is required. The wallet, called `wallet`, has a default value which you are to set to the largest amount of money needed if all `numBets` are placed.  This function returns gain. Note that your code should check that `numBets` and `wallet` are compatible, i.e., that there is enough money in the wallet for the number of bets requested. If not, then lower the number of bets to a value that matches the size of the wallet and issue a warning that explains the problem and provides the revised number of bets. (To do this you can `paste` text together with objects, e.g., 

```{r}
x = 2
paste("The value of x is ", x, "Were you surprised?")
```


Write your function in the code chunk below


```{r, error = TRUE}
double.both = function(numBets, wallet = 2^numBets-1) {
  if (floor(sqrt(wallet+1)) < numBets){
    warning (paste("The amount of money you have is ", wallet, "Were you surprised?"))
    numBets = floor(sqrt(wallet+1))
  }
  return(double.revised(numBets))
}
```

Test your function
```{r, collapse = TRUE, error=TRUE}
double.both(4)
double.both(4, 1)
double.both(1, 4)
double.both(10, 100000)
```

