---
title: "Lab 6: Regular Expression Practice"
output: html_document
---

## Introduction

In this lab, you will write some regular expressions to extract features from text data. We will use a small sample of the tweets that you will use for your homework.

These are available on the Web. Load them into your session with

```{r}
load(url("http://www.stat.berkeley.edu/users/nolan/data/testTweets.rda"))
```

Explore the data frame. How many variables does it have, what's the class of each, and how many tweets are included?
It has 3 variables. 
Ttext       Tdate        Tdev 
"character" "character" "character" 
It has 5 tweets

```{r}
str(testTweets)
```

## Retweets

The text of the tweet is in the variable `Ttext` in the data frame. Some of these tweets are retweets. You can tell that they are retweets because the begin with an RT. Create a logical vector that is TRUE if the tweet is a retweet. Call the vector `RT`:

```{r error=TRUE}
RT = grepl("^RT",testTweets$Ttext)
```

## When the tweet was sent

We are interested in the hour of the day that the tweet was sent. The day, hour, minute, second, and year are 
provided in `Tdate`.

Use regular expressions to extract the hour of the day that the tweet was sent. Make sure that you save the hour in a numeric vector called `hr`.

```{r, error = TRUE}
hr = as.numeric(sapply(testTweets$Tdate, function(s) substr(s, 9, 10)))
```

## Hash Tags

Can you find which tweets contain a hashtag? Hashtags begin with #, and do not include punctuation or blanks. Create a logical vector that is `TRUE` if a tweet contains 1 or more hashtags. Be sure that it doesn't consider a "#" on its own as a hash tag. Call this logical vector `anyHash`

```{r, error=TRUE}
anyHash = grepl("#[^[:space:]][^[:punct:]]",testTweets$Ttext)
mytext = c("# ", "#loevlyday", "#,", "#b,", "#b A")
grepl("#[^[:space:]][^[:punct:]]",mytext)
```

The `anyHash` logical should be 
```{r, eval=FALSE}
anyHash
[1] FALSE  TRUE  TRUE  TRUE FALSE
```


Use `gregexpr` to find the starting and ending point of the hash tags in the tweets that have a hash tag. Check that you correctly identify the starting point and length of the match. For example, one tweet contains "...  #MAGA. ..." and you do not want to include the period as part of the tag.
Also, be sure that your call finds all of the hashtags present in the tweet (some tweets have 2 or 3). Save this information in `hashLocs` 

```{r, error=TRUE}
hashLocs = gregexpr(pattern = "#[A-z]*",
                    testTweets$Ttext)
hashLocs
```

This call to `gregexpr` returns a list. The 1st and 5th elements should be a vector consisting of -1.
The third element is a numeric vector of length 2 with values 71 and 87 and an attribute called `match.length` of 15 and 8. 

We can use `hashLocs` to extract the hash tags with `substr`. The following code does that for us.

```{r, error=TRUE}
hashtagList = 
  mapply(function(x, sta) {
    substr(rep(x, length(sta)), 
           start = sta, 
           stop = sta + attributes(sta)$match.length - 1)
  },
         x = testTweets$Ttext,
         sta = hashLocs)
names(hashtagList) = NULL
hashtagList
```

The `hashtagList` is a list of 5-elements. The first and fifth are empty strings. The second is "#MAGA", and the 3rd and 4th each have 2 hashtags:  "#BigLeagueTruth" and "#Debates" for the 3rd, and "#TrumpTrain" and "#Debates2016" for the 4th.
