---
title: "IMDB - Getting Started"
output: html_document
---

To connect to the database, 

```{r}
library(RMySQL)
drv = dbDriver("MySQL")
con = dbConnect(drv, dbname = 'imdb', user = 'imdb', password = 'imdb',
                host = 'radagast.berkeley.edu')
dbListTables(con)
```

Answer the following 2 questions. 

## How many records are in the `kind_type` table? 

Put your answer in `numKind`. Retrieve all of the records and display print them to the console.

```{r}
dbGetQuery(con, "SELECT * FROM kind_type;")
numKind = dbGetQuery(con, "SELECT COUNT(*) FROM kind_type;")
numKind
```

## How many movies and TV series are in imdb?

Use the information in `kind_type` to determine how many television series are in the data base and how many movies are in the data base. Put your answers in `numTV` and `numMov`, respectively

```{r}
dbListFields(con, "title")
numTV = dbGetQuery(con, "SELECT COUNT(*) FROM title WHERE kind_id = 2;")
numMov = dbGetQuery(con, "SELECT COUNT(*) FROM title WHERE kind_id = 1;")
```

## What is Brad Pitt's id?

Pull all of the names of actors and their ids  into R. Use regular expressions to find the identification number for Brad Pitt.  Place this id number in `PittID`

```{r}
dbListFields(con, "char_name")
name_id = dbGetQuery(con, "SELECT id, name FROM name;")
filter = grepl("Pitt, Brad", name_id$name)
PittID = name_id$id[try]
```

## Additional Information

Examine all of the fields for the first 10 records in the `info_type` table.
If you wanted to find genre information out about a movie, which id would you use?  Find the title and production year for 10 movies in the comedy genre. 

```{r}
dbListFields(con, "info_type")
first_10 = dbGetQuery(con, "SELECT * FROM info_type limit 10;")
genre = 3
dbGetQuery(con, "SELECT t.title, t.production_year FROM title AS t
                 INNER JOIN movie_info on t.id = movie_info.movie_id
                 WHERE info_type_id = 3  and info = 'Comedy' 
                       LIMIT 10;")
```

