---
title: "Project Ideas"
output: html_document
---

## Bounding Box for Node Density

You will be generating data in a 3-dimensional rectangle that encompasses the population density. We know the (x, y) space is 100 by 100, but we need to figure out the height in the z dimension. To do this generate a dense grid in the (x, y) plane.  Make a vector of x's and a vector of y's that are 200 long and range from 0 to 100.

```{r error = TRUE}
x = seq(from = 0, to = 100, length.out = 200)
y = seq(from = 0, to = 100, length.out = 200)
```


Now use `expand.grid()` to create a grid of all combinations of x and y

```{r, error=TRUE}
gridPts = expand.grid(x, y)
```

Check the dimensions of gridPts. It should have 200 * 200 rows.

```{r, error=TRUE}
dim(gridPts)
```

Pass `gridPts` to `nodeDensity` and find the maximum value returned.
Use this value plus a smidgen as the hieght of the 3-dimensional cube.
Call is `zMax`.

```{r}
source("http://www.stat.berkeley.edu/users/nolan/data/nodeDensity.R")
zMax = ceiling(max(nodeDensity(gridPts$Var1, gridPts$Var2)))
```


## Generating Points in the 100 by 100 by zMax box

Use `runif` to generate 200 points in the 3-dimensional region

```{r, error = TRUE}
x200 = runif(200, min = 0, max = 100)
y200 = runif(200, min = 0, max = 100)
z200 = runif(200, min = 0, max = zMax)
```

Figure out which points have a z-value under the population density function.  Use a logical expression to do this

```{r}
keep = z200[nodeDensity(x200, y200) > z200]
```


How many points are kept?
```{r}
sum(keep)
```

If half the points are kept then you probably want to generate twice as many (or more) in order to get enough. 

You will want to develop this code into the `genNodes` function.
Note that to be sure that you return `n` (x,y) points (no more and no less) you will want to have a mechanism to continue to generate points until you have `n` or more, and then return `n` of them. 
